---
title: "Developer Notes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3) Developer Notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(zctaCrosswalk)
library(dplyr)
```

While creating this package I was acutely aware that ZCTAs change frequently. For example,
back in 2016 I created the [choroplethrZip](https://github.com/arilamstein/choroplethrZip) package. That package is now out of date, 
because the underlying data it stores became out of date. I expect that something similar 
will eventually happen with this package.

This vignette is written as 
a "note to my future self" in case I wind up needing to write a similar package 
again in the future. It is also intended to increase the number of people who 
understand how to create 
packages like this. 

## ?zcta_crosswalk
The core data structure in this package is `?zcta_crosswalk`:

```{r}
data(zcta_crosswalk)

print(zcta_crosswalk, n = 5)
```

Let's examine how the data structure was created.

### ?get_zcta_crosswalk

Start by looking at the contents of the function `?get_zcta_crosswalk`. You can
do this by typing `get_zcta_crosswalk` at the console:

```{r}
# By not including the trailing (), R will show you the function's contents
get_zcta_crosswalk
```

The function starts by reading and transforming the contents of a URL. At the 
time of this writing
that is the URL for the Census Bureau's "2010 ZCTA to County Relationship File", 
a file which I mentioned elsewhere.

This means that if (for example) Census publishes new data in the same format tomorrow, 
you could just change the URL, rerun the same code and get the updated data in R. 
(Note 
that I do not know when Census plans to update this dataset or whether they 
plan to publish it in the same format.)

If you open the URL referenced in `?get_zcta_crosswalk` in a browser you will see rows like this:

```
221704258470394|90210|ZCTA5 90210|27823432|153478|G6350|B5|S|275901063468976|06037|Los Angeles County|10513491099|1787501506|G4020|H1|A|27823432|153478
```
This tells us that ZCTA 94109 is in Los Angeles County. It also tells us that Los 
Angles County has FIPS Code 06037.

### Adding in State Information

Unfortunately, the file does not directly contain any state information. And 
since I wanted to run queries like "Get all ZCTAs in a given state", I needed 
to add that in. 

I started by splitting out the first two characters of a County FIPS Code into 
a new column called `state_fips`. This allows a user to search for ZCTAs in a state if they know the state's FIPS code. 

However, this does not help us if users want to select identify a state by it's name or Postal Code Abbreviation.
To address this limitation, I created a new dataframe called `state_names`, and use it to join against the results of `get_zcta_crosswalk`:

```{r}
data(state_names)

print(state_names, n = 5)
```

One thing to keep in mind is that while there are technically only 50 "states", "state" in this dataset really means "any top level administrative region". This dataset has 56 "states" (the "extra" ones are the District of Columbia, Puerto Rico, US Virigin Islands, American Samoa, Guam and the Northern Mariana Islands).

I believe that it would be useful for R to have a standalone package that contains 
a data frame like this for all FIPS codes. I did not do that here because even 
though this dataset has 56 state-level entities, the full list of FIPS codes is 
much larger (e.g. see [here](https://www.census.gov/library/reference/code-lists/ansi.html)).

Also note that while R has two built-in vectors that deal with state names (`state.abb` and `state.name`) but they cannot help us for two reasons: (1) they do not contain FIPS code and (2) they only contain 50 states.

## Learning About ZCTAs

If you would like to learn more about ZCTAs (including how they differ from ZIP 
Codes), I recommend two references:

1. My free course [Mapmaking in R with Choroplethr](https://ari-lamsteins-courses.thinkific.com/courses/mapmaking-in-r-with-choroplethr) has three sections dedicated to "ZIP Code Choropleths".
2. In 2017 I had the pleasure of meeting Jon Sperling, who is one of the creators of the ZCTA, at the Association of Public Data Users (APDU) conference. You can learn about that meeting, including a reference to one of his papers on the topic, [here](https://arilamstein.com/blog/2017/10/24/meeting-titans-open-data/). 

One of my recollections from that meeting is Jon explaining that ZIP Codes are designed to follow roads. This means that different sides of the same block can have different ZIP codes.
Census geography, however, is built on 
"blocks". This means that all homes on an entire block must be the same ZCTA. This 
difference in construction means that ZIPs and ZCTAs are unlikely to ever truly be identical.

Some other details that might interest people new to the field:

 * Unlike States and Counties, ZCTAs do not "cover" the entire country. The full ZCTA map contains "holes".
 * ZCTAs frequently cross counties and sometimes even states. These cases appear as duplicate entries in `zcta_crosswalk`. (For example, six rows have ZCTA 39573).
 * The primary reason why `get_zctas_by_county` does not accept county names is that county names are not unique between states. (For example, 30 states have a "Washington County").

## Funding

I would like to thank my employer, [MarketBridge](https://market-bridge.com/), for supporting my development of this package. This package would not have been developed without their support.
