---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(zctaCrosswalk)
```

This package is designed to help answer common analytical questions that arise
when working with US ZIP Codes.

Note: the entity which maintains US ZIP Codes (the US Postal
Service) does not release a map or crosswalk of that dataset. As a result, most
analysts instead use [ZIP Code Tabulation Areas (ZCTAs)](https://www.census.gov/programs-surveys/geography/guidance/geo-areas/zctas.html) which
are maintained by the US Census Bureau. Census
also provides [Relationship Files](https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.2020.html#zcta) that maps ZCTAs to other geographies.

This package provides the "2020 ZCTA to County Relationship File" as a tibble, combines it with useful publicly available metadata (such as State names) and provides convenience functions for querying it.

The most useful functions and data objects in this package are:

 * `?get_zctas_by_state`
 * `?get_zctas_by_county`
 * `?get_zcta_metadata`

## ?get_zctas_by_state

`?get_zctas_by_state` takes a vector of states and returns the vector of ZCTAs in those states. Here are some examples:

```{r}
# Not case sensitive when using state names
head(
  get_zctas_by_state("California")
)

# USPS state abbreviations are also OK - but these *are* case sensitive
head(
  get_zctas_by_state("CA")
)

# Multiple states at the same time are also OK
head(
  get_zctas_by_state(c("CA", "NY"))
)

# Throws an error - you can't mix types in a single request
# get_zctas_by_state(c("California", "NY"))
```

Note that this dataset's conception of "State" is likely different than yours. 
The primary data structure in this package is `?zcta_crosswalk` and contains 
56 "states":

```{r}
data(zcta_crosswalk)

head(
  unique(zcta_crosswalk$state_name)
)
```

A common problem when doing analytics with states is ambiguity around names. For 
example, most people write "Washington, DC". But this dataset uses "District of Columbia". The most common solution to this problem is to use [FIPS Codes](https://en.wikipedia.org/wiki/Federal_Information_Processing_Standard_state_code)
when doing analytics with states. And so `?get_zctas_by_state` also
supports FIPS codes. 

Note that technically FIPS codes are characters and have a leading 
zero (e.g. California is "06"). But in practice people often use numbers (e.g.
6 for California) as well. As a result, `?get_zctas_by_state` supports both:

```{r}
ca1 = get_zctas_by_state("CA")
ca2 = get_zctas_by_state("06")
ca3 = get_zctas_by_state(6)
all(ca1 == ca2)
all(ca2 == ca3)
```

## ?get_zctas_by_county

`?get_zctas_by_county` works analogously to `?get_zctas_by_state`. The primary
difference is that it only accepts FIPS codes. This is because [FIPS county codes](https://en.wikipedia.org/wiki/FIPS_county_code) are unique, but their names are not.
(For example, 30 counties in this dataset are named "Washington County"!)

If you need to find the FIPS code for a particular county, I recommend simply googling
it (e.g. "FIPS code for San Francisco County California") or consulting
[this](https://en.wikipedia.org/wiki/List_of_United_States_FIPS_codes_by_county) page.

Note that the FIPS codes can be either character or numeric.
```{r}
# "06075" is San Francisco County, California
head(
  get_zctas_by_county("06075")
)

# 6075 (== as.numeric("06075")) works too
head(
  get_zctas_by_county(6075)
)

# Multiple counties at the same time are also OK
head(
  get_zctas_by_county(c("06075", "36059"))
)
```

## ?get_zcta_metadata

`?get_zcta_metadata` takes a vector of ZCTAs and returns all available metadata on them.
The ZCTAs can be either character or numeric.

```{r}
get_zcta_metadata("90210")

get_zcta_metadata(c(90210, 90211))
```
